\large Fowizzle: Combining spectra or images by forward modeling

### 2023-11-02 paper/main.tex
Updates from Overleaf
start de-escalation of the language

% Copyright 2022 David W. Hogg (NYU) and Andy Casey (Monash)

% To-Do items
% -----------
% - Audit for rigid terminology. HOGG is currently liking ``combined spectrum'' for the model expectation, for example. And "standard practice" for the traditions of our field.
% - Audit and fix the sign applied to $\Delta x_i$ everywhere. Wrong relative to code?
% - Write about the two input-data cases.
% - See bug list in code for other issues with the text--code relationship (noise model; line-formation model).
% - Audit for its / it's. Hogg has issues with this.
% - Deal with all HOGG, ANDY, ARC, XXX, YYY, CITE, etc.

@@

\title{\Huge Don't interpolate the data!}
% ARC: I changed this to "Don't interpolate the data!"? because you can't like, own data, man.
% (but you can change it back if you prefer)

@@

\begin{abstract}\noindent
When there are many observations of an astronomical source---many images with different dithers, or many spectra taken at different barycentric velocities---it is often tempting to shift and stack the data, to (for example) make a high signal-to-noise average image or mean spectrum.
Bound-saturating measurements are made not by manipulating data, but instead by manipulating a likelihood function, where the data are treated as fixed, and model parameters are modified to fit the data.
Traditional shifting and stacking of data can be converted into a model-fitting procedure, such that no data are ever harmed, and yet the output is the shift-adjusted mean.
The key component of this conversion is a spectral model that is completely flexible but also a continuous function of wavelength (or position in the case of imaging) that can represent any signal being measured by the device after any reasonable translation (or rotation or field distortion).
The benefits of a modeling approach are myriad:
The sacred data never are modified.
Noise maps, data gaps, and bad-data masks don't require interpolation.
The output can take the form of a spectrum evaluated on a pixel grid, as is traditional.
The noise in the output spectrum becomes uncorrelated across neighboring pixels. 
The only cost is a small increase in computational complexity.
We demonstrate all of these things with a small data example and we provide open-source sample code for re-use.
\end{abstract}

@@

It would be wrong to say that the practice of astronomy is, basically, \emph{staring at the sky}.
That said, staring at the sky is a big part of what we do.
And the way we do it, usually, is this:
We take many individual images of our target with exposure times that are long enough to make read noise irrelevant (if possible) while at the same time short enough to avoid saturation (of what we care about).
We then shift and average (maybe median?) those images to get the highest possible signal-to-noise on our target.
This, fundamentally, is the technology underlying the \textsl{Hubble Deep Field} (\citealt{hdf}),
the \textsl{APOGEE} spectroscopic survey (\citealt{apogee}),
and the future Rubin Observatory \textsl{LSST} (\citealt{lsst}), among countless other projects.

Is it astronomical blasphemy to say that this practice is \emph{wrong?}
It is wrong because minimum-variance unbiased estimators require care in their construction; they are (often) maximum-likelihood estimators (\citealt{something}). % HOGG
The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of the image or spectrum, given the noisy data of the individual exposures.
Noisy data are combined to make a measurement by building and optimizing a likelihood, not by shifting and stacking the input data.
Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator}, but it is.
It is just a simultaneous estimate of a lot of different components (a lot of different pixels; all the pixels in the combined image or spectrum).

@@

One of the great things about the standard practice---the simple process of shifting and stacking the data---is that this procedure makes almost no assumptions about what the output, combined image or spectrum will look like.
It isn't really a ``model'' for the data; it is more like a ``summary'' of the data.
When we switch to a statistical-inference framework, we have to choose some form or basis for expressing the combined image.
In what follows we use a maximally parameterized model, in the sense that we choose an expression for the functions or function space that has as many parameters as the output image has pixels.
That means that the approach we take does not get its power from the choice of parameterized functions for the output:
Indeed it has just as much freedom as the shifting-and-stacking standard practice.
The forward-modeling we recommend gets its power from the fact that it generates the data directly and without manipulation.

The data are never manipulated when using an estimator (in contrast to standard practice), but (for us) the functional requirements are the same as with standard practice:
We consider here only procedures that start with the individual, single-epoch observations as inputs, along with noise estimates, bad-pixel masks, and perhaps bit-flags; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates and metadata flags.
That is, the only difference between standard practice and the forward model that we advocate is the purely in the nature of the estimator.
Unlike the shift-and-coadd approach, the estimator we will deliver will be insensitive to under- or poorly-sampled data, which is not uncommon when astronomical seeing accidentally gets excellent, or in spectrographs that have limited detector pixels (for example, \citealt{apogeehardware}).
As we will show, an estimator also often yields a more faithful estimate of the \emph{true} spectrum, and almost always better noise properties.
Both of these results make our results better for any subsequent precision measurements (e.g., radial velocities, chemical abundances).
But importantly, we aren't proposing a change to the \emph{format} of the input or outputs of spectral or image combination.

>>>>

% Copyright 2022, 2023, [add your years here] David W. Hogg (NYU) and Andy Casey (Monash)

% To-Do items
% -----------
% - Change the tone to "here's a method; it is principled and has great properties" and remove all suggestion that it is THE BEST THING.
% - Put in notes about our hypothesis for WHY we get uncorrelated pixels.
% - Make comments directly about Drizzle to justify the name.
% - discuss conservation of flux vs intensity.
% - Blanton asks: Is this identical to sinc shift and average? Hogg thinks it MIGHT be.
% - Add Hogg response to Blanton: Interpolate with (say) cubic spline and then average is NOT THE SAME as fit all the data simultaneously with a cubic spline function on a particular set of knots.
% - Dustin asks: Is this effectively producing a sinc LSF? Hogg thinks it DEFINITELY IS NOT. Explain why not in the text.
% - Hogg notes: Is it clear in the text that we are taking everything to a standard instrument? And the user can decide what that instrument is? That isn't in our current implementation, but it COULD BE. Discussion point?
% - Related Hogg issue: Any latent spectrum (like our combined spectrum) should be something that COULD BE OBSERVED by some instrument in the future. Hence spectro-perfectionism is wrong (ish).
% - Read and cite the DESI spectral pipeline paper; Dustin says it's relevant.
% - Audit for rigid terminology. HOGG is currently liking ``combined spectrum'' for the model expectation, for example. And "standard practice" for the traditions of our field.
% - Audit and fix the sign applied to $\Delta x_i$ everywhere. Wrong relative to code?
% - Write about the two input-data cases.
% - See bug list in code for other issues with the text--code relationship (noise model; line-formation model).
% - Audit for its / it's. Hogg has issues with this.
% - Deal with all HOGG, ANDY, ARC, XXX, YYY, CITE, etc.

@@

\title{\large Fowizzle: Combining spectra or images by forward modeling}

@@

\begin{abstract}\noindent
When there are many observations of an astronomical source---many images with different dithers, or many spectra taken at different barycentric velocities---it is standard practice to shift and stack the data, to (for example) make a high signal-to-noise average image or mean spectrum.
Bound-saturating measurements are made by manipulating a likelihood function, where the data are treated as fixed, and model parameters are modified to fit the data.
Traditional shifting and stacking of data can be converted into a model-fitting procedure, such that the data are not modified, and yet the output is the shift-adjusted mean.
The key component of this conversion is a spectral model that is completely flexible but also a continuous function of wavelength (or position in the case of imaging) that can represent any signal being measured by the device after any reasonable translation (or rotation or field distortion).
The benefits of a modeling approach are myriad:
The sacred data never are modified.
Noise maps, data gaps, and bad-data masks don't require interpolation.
The output can take the form of an image or spectrum evaluated on a pixel grid, as is traditional.
In addition to shifts, the model can account for line-spread or point-spread function variations, world-coordinate-system variations, and calibration or normalization variations.
The noise in the output becomes uncorrelated across neighboring pixels as the shifts deliver good coverage in some sense.
The only cost is a small increase in computational complexity.
We demonstrate the method with a small data example and we provide open-source sample code for re-use.
\end{abstract}

@@

It would be wrong to say that the practice of astronomy is, basically, \emph{staring at the sky}.
That said, staring at the sky is a big part of what we do.
And the way we do it, usually, is this:
We take many individual images of our target with exposure times that are long enough to make read noise irrelevant (if possible) while at the same time short enough to avoid saturation (of what we care about).
We then shift and average (maybe median?) those images to get the highest possible signal-to-noise on our target.
This, fundamentally, is the technology underlying the \textsl{Hubble Deep Field} (\citealt{hdf}),
the \textsl{APOGEE} spectroscopic survey (\citealt{apogee}),
and the future Rubin Observatory \textsl{LSST} (\citealt{lsst}), among countless other projects.
This technology is encoded in tools such as \textsl{Drizzle} (\citealt{drizzle}) and HOGG.

These traditional practices are reliable and fast.
However, they will---in some cases---sacrifice some information.
Minimum-variance unbiased estimators require care in their construction; they are generally maximum-likelihood estimators (\citealt{something}). % HOGG
The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of some kind of ``true'' or high signal-to-noise image or spectrum, given the noisy data of the individual exposures.
That is, a combined image is a kind of \emph{prediction}: It is the prediction of what \emph{would have been observed} with a much longer exposure with a much better device.
In measurement theory, the practice is to combine noisy data into a measurement by building and optimizing a likelihood, not by shifting and stacking (or performing other arithmetic operations on) the input data.
Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator}, but it is.
It is just a simultaneous estimate of a lot of different components (a lot of different pixels; all the pixels in the combined image or spectrum).

@@

One of the great things about the standard practice---the simple process of shifting and stacking the data---is that this procedure makes almost no assumptions about what the output, combined image or spectrum will look like.
It isn't really a ``model'' for the data; it is more like a ``summary'' of the data.
When we switch to a statistical-inference framework, we have to choose some form or basis for expressing the combined image.
In what follows we use a maximally parameterized model, in the sense that we choose an expression for the functions or function space that has as many parameters as the output image has pixels.
That means that the approach we take does not get its power from the choice of parameterized functions for the output:
Indeed it has just as much freedom as the shifting-and-stacking standard practice.
The forward-modeling we recommend gets its power from the fact that it generates the data directly and without manipulation, by means of a justified likelihood function.

A boundary condition of the work presented here is that the functional requirements of anything we provide must be the same as with standard practice:
That is, we consider here only procedures that start with individual, single-epoch observations as inputs, along with noise estimates, bad-pixel masks, and perhaps bit-flags; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates and metadata flags.
That is, the only difference between standard practice and the forward model that we advocate is the purely in the nature of the estimator.
Unlike the shift-and-coadd approach, the estimator we will deliver will be insensitive to under- or poorly-sampled data, which is not uncommon when astronomical seeing accidentally gets excellent, or in spectrographs that have limited detector pixels (for example, \citealt{apogeehardware}).
As we will show, an estimator also often yields a more faithful estimate of the \emph{true} spectrum, and almost always better noise properties.
Both of these results make our results better for any subsequent precision measurements (e.g., radial velocities, chemical abundances).
But importantly, we aren't proposing a change to the \emph{format} of the input or outputs of spectral or image combination.
